%================================================================================
%       Safety Critical Systems Club - Data Safety Initiative Working Group
%================================================================================
%                       DDDD    SSSS  IIIII  W   W   GGGG
%                       D   D  S        I    W   W  G   
%                       D   D   SSS     I    W W W  G  GG
%                       D   D      S    I    WW WW  G   G
%                       DDDD   SSSS   IIIII  W   W   GGG
%================================================================================
%               Data Safety Guidance Document - LaTeX Source File
%================================================================================
%
% Description:
%   Data Safety Management Plan section.
%
%================================================================================
\section{\ecapitalisewords{\glsentrylong{dsmp}} (Informative)} \label{bkm:plan}

\dsiwgSectionQuote{Things get done only if the data we gather can inform and inspire those in a position to make a difference.}{Mike Schmoker}

This section gives a suggested \gls{dsmp} table of contents. It is expected that this will be needed only for aspects not already covered in a \gls{smp}, or similar. It can be merged with an \gls{smp}, if appropriate. However it may be useful to consider the distinct data perspective by using a \gls{dsmp} as well as an \gls{smp}. Regardless, a close connection should be maintained between the \gls{smp} and the \gls{dsmp}.

\Gls{dsmp} suggested contents:
\begin{enumerate}
  \item Introduction:
  \begin{description}
    \item[Scope and context:] sets the scene, describes the project, scenario, concept of operations, etc.;
    \item[Boundaries and interfaces:] Describes the main interfaces and exchanges, with a scope boundary diagram.;
    \item[Derived requirements:] System-level requirements which impact the data safety process;
    \item[Owners:]\index{Data!Owner} Who owns the data under consideration as it progresses through the system?;
    \item[Producers / consumers:] Who are the producers and consumers of the data the system inputs and outputs?;
    \item[Assumptions];
    \item[References]; and
    \item[Abbreviations and acronyms].
  \end{description}
  \item Analysis of Assigned \gls{dsal}\index{Assurance Level!Data} and \gls{odr} Level (Implications of the data analyses.):
  \begin{description}
    \item[\Gls{sil}, etc., implications:] What impact does the \gls{dsal}\index{Assurance Level!Data} have on the required \gls{sil}, or similar measure, and vice versa?;
    \item[Development implications:] Are there any special development considerations? Derived from the \gls{sil} if there is one, otherwise what is deemed necessary for this system.;
    \item[\Gls{verification} implications:] derived from the \gls{sil} if there is one, otherwise what is deemed necessary for this system.;
    \item[Assurance implications:] derived from the \gls{sil} if there is one, otherwise what is deemed necessary for this system.; and
    \item[Process / procedure implications:] derived from the \gls{sil} if there is one, otherwise what is deemed necessary for this system.
  \end{description}
\item
  Categories
  of safety data in scope (a list of all the categories\index{Category!Data} to be considered in the system context.).
  \item Data requirements analysis:
  \begin{description}
    \item[Lifecycles:]\index{Lifecycle!Data} what data lifecycles are to be used?;
    \item[Specific Targets:] are there any qualitative or quantitative targets for the data?; and
    \item[Security Considerations:] how will security be managed in this context? Are there any security / safety conflicts? Are there any security-related causes of data \glspl{hazard}?
  \end{description}
  \item Management approach (how will the organization manage the data safety risks?):
  \begin{itemize}
    \item Organization;
    \item Responsibilities;
    \item Authorisations; and
    \item Approvals and signoffs.
  \end{itemize}
  \item Justification Approach (how will the safe use of the data be justified, e.g., as part of the safety case report?). 
  \item Analyses / \glspl{verification} to be performed (what analyses or checks are to be performed on the data?).
  \item Documents to be produced (The list of documents to be produced related to data aspects.).
  \item Appendix: \gls{dsal}\index{Assurance Level!Data} guidelines response (tailored version of the tables from this document. What is considered applicable / useful and what is not?).
\end{enumerate}
